---
title: "Insurance Charges"
subtitle: "Does Geography Matter?"
author: "Cabrina S."
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    self_contained: false
---

## 1. Introduction

When we look at medical insurance premiums, we often wonder if moving to a different part of the country could change our medical premiums (bills).

**Our Investigation:** Today we are using a dataset of 1,338 insurance customers to see if there is a statistically significant relationship between *Region* and *Medical Charges*.

**The Roadmap:**

* Visualize the "Spread of Costs"
* Define our statistical "Ground Rules"
* Testing for significant differences using ANOVA
* Understanding how health factors (BMI/AGE) complicate the story  


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning= F,message= F)
library(readr)
library(plotly)
library(ggplot2)
library(dplyr)

ds <- read.csv("insurance.csv")
```

## Summary of the data we are using

Before diving into the analysis, we verified the data for 1,338 individuals to ensure there were no illogical extremes, like an age of 200.
```{r}
ds %>% select(age, bmi, charges) %>% summary()
```

## First, Does one region have a higher mean?


```{r Outliers_and_NO_Outliers, fig.height=3.5, fig.width=5}

ggplot(ds, aes(x = region, y = charges, fill = region)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Medical Charges by Region~With Outliers",
       x = "US Region", y = "Charges ($)") +
  scale_fill_brewer(palette = "Set2")+ 
  theme(legend.position = "none")


ggplot(ds, aes(x = region, y = charges, fill = region)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  coord_cartesian(ylim = c(0, 20000)) + 
  theme_minimal() +
  labs(title = "Medical Charges by Region~Without Outliers",
       x = "US Region", y = "Charges ($)") +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")

```


A boxplot is the best way to see the "spread" and "outliers" of costs for each region.  
By removing the outliers in the second graph, we can focus on charges that are under $20,000.  
We can see that the median costs are actually quite similar, though the Northeast trends slightly higher with the Southeast right behind it.

## Second, Where is the Money?

```{r Density_by_region}
ggplot(ds, aes(x = region, y = charges, fill = region)) +
  geom_violin(alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.1, color="black") +
  theme_classic() +
  labs(title = "Distribution of Medical Charges by Region", subtitle = "Wider areas show where most people's costs fall. Southeast shows higher 'peaks'.",
       x = "US Region", y = "Insurance Charges ($)")+
  theme(legend.position = "none")
```
<br>
This violin plot shows us where the "density" of people lies.  
Despite the southeast having a lesser amount of people, the taller point shows that they have a higher density of expensive claims.


## The Logic of ANOVA test

To prove that region matters, we can't just look at averages; we have to look at **Variance**. We use a test called **ANOVA**.  

The **F-statistic** is calculated as:

$$F = \frac{\text{Variance between regions}}{\text{Variance within regions}}$$  
Where:  
**Variance between regions:** is how much the regional averages differ from each other.  
**Variance within regions:** is how much the individuals vary within their own region.  

If the variance **between** regions is much larger than the variance **within** them, we conclude that the location has a significant impact on charges.  


## Setting Ground Rules

For our conclusions to be reliable, our data must follow three main rules:  

1. **Independence:** One person's bill does not affect anothers.
2. **Normality:** The variation in our data follows a bell curve:
   $$\epsilon \sim N(0, \sigma^2)$$
3. **Equal Variance:** The "spread" of costs is roughly similar across all regions:
   $$\sigma^2_{NE} \approx \sigma^2_{NW} \approx \sigma^2_{SE} \approx \sigma^2_{SW}$$

If these hold, our $p\text{-value}$ will accurately tell us if Region is a significant factor.

## Test {.smaller}

Now we calculate our p-value. If the p-value is less than 0.05, we reject the idea that all regions are equal.   
We do that with this code:

```{r ANOVA_test,echo=TRUE,results= 'asis'}

region_model <- aov(charges ~ region, data = ds)
anova_summary <- summary(region_model)

p_value <- anova_summary[[1]]["region", "Pr(>F)"]

cat("ANOVA Result\n")
cat(sprintf("**p-value = %.4f**\n\n", p_value))
```

```{r}
if (p_value < 0.05) {
  cat("- **Statistically significant difference between regions**")
} else {
  cat("+ **No statistically significant difference between regions**")
}
```

## The "Hidden" Variables: Age and BMI

There is a positive linear trend between BMI and medical charges. Most importantly, the trend lines for all four regions are nearly parallel. This tells us that while a region might be slightly more expensive overall, the higher BMI is a better indicator for increased charges.

```{r Age_BMI}
ggplot(ds, aes(x=bmi, y= charges, color=region))+
  geom_point(alpha= 0.4)+
  geom_smooth(method= "lm", se=F)+
  theme_minimal()+
  labs(title= "Charges VS. BMI across Regions")
```


## The Big Picture: 3D Interactive Graph Code  


Here we have the code for an interactive model that maps 
* 'Age' on the x-axis
* 'BMI' on the y-axis
* 'Chagres ($)' on the z-axis.   
```{r Age_BMI_byRegn, echo=T, eval=FALSE}
plot_ly(ds, x = ~age, y = ~bmi, z = ~charges, 
        color = ~region, colors = "Set1",
        type = "scatter3d", mode = "markers", marker = list(size = 3)) %>%
  layout(title = "The Complexity of Costs",
         scene = list(xaxis = list(title = 'Age'),
                      yaxis = list(title = 'BMI'),
                      zaxis = list(title = 'Charges ($)')))
```

## The Big Picture: 3D Interactive Graph

```{r Age_BMI_Charge_graph}
plot_ly(ds, x = ~age, y = ~bmi, z = ~charges, 
        color = ~region, colors = "Set1",
        type = "scatter3d", mode = "markers", marker = list(size = 3)) %>%
  layout(title = "The Complexity of Costs",
         scene = list(xaxis = list(title = 'Age'),
                      yaxis = list(title = 'BMI'),
                      zaxis = list(title = 'Charges ($)')))
```
Costs aren't just about region; they are driven by health. Here we see how Age and BMI drive Charges, colored by Region.

## Final Verdict: Does Region Matter?

### **The Findings:**
Our ANOVA test (Slide 5) typically shoes a low p-value, meaning region *does* influence cost.  
The Southeast region often has the highest average charges.  
<br>
**The 'But':**  
While geography matters, our 3D and BMI plot shows that *Age & BMI* create the largest cost differences.  
<br>
**Conclusion:** Where you live matters for your insurance bill, but who you are and your health lifestyle matter more.